An item response theory analysis of the matrix reasoning item bank (MaRs-IB)

### 2022-01-16 manuscript/main.tex
writing day writing day

Despite their utility, there are notable challenges in using matrix reasoning measures. The first is that of copyright. The most prominent measures, including the Raven's progressive matrices \citep{raven2003raven} and the Wechsler Abbreviated Scale of Intelligence matrix reasoning subtest \citep{wechsler1999wechsler}, are not free to use and copyright may prevent these pen-and-paper tasks from being adapted into computerized tasks. A second challenge is that measures that are free-to-use and in the public domain typically have very few items available. For example, the Hagen Matrices Test \citep{heydasch2014hagen} and International Cognitive Ability Resource \citep{condon2014international} have only 20 and 16 items available, respectively. The limited number of items raises the issue of practice effects \citep{ng1974applicability, bors2003effect}, especially in the context of online experiments where many lab groups may be testing the same participants.

To address these gaps, \cite{chierchia2019matrix} developed the matrix reasoning item bank (MaRs-IB). The MaRs-IB is an open-source, free-to-use set of 80 matrix reasoning items, each with three unique shape variants and two sets of distractors. The items span a wide range of complexity -- in terms of both their number of features and number of changing relations. Example items here?  In a sample of over 600 adults, adolescents, and children, the authors found that a 8-minute MaRs-IB test measure had excellent psychometric properties (high internal consistency and test-retest reliability) and convergent validity (correlations with ICAR matrix reasoning and digit span performance). Most importantly, the authors released item-level performance data for each item, thereby raising the possibility for other researchers to yield age-appropriate tasks of custom difficulty and duration.

There are, however, two impediments in researchers using the MaRs-IB norms. Both concern the design of the original MaRs-IB task. The initial MaRs-IB task was a fixed-time task, wherein participants completed as many items as possible in 8 minutes. The first issue then is that, because participants completed 33 items on average, there exists relatively little data on the majority of the MaRs-IB items (this is before taking into consideration the data available for each item variant). The second issue is that, because of individual differences in speed-accuracy trade-offs, the estimates of item difficulties may be biased. That is, items completed by fewer participants are more likely to have been reached by participants who had been sacrificing accuracy for speed, thereby giving a distorted view of the difficulty of these items.

Summary paragraph: here we test a large sample of adults, making sure that all items and variants were tested many times. We fix the number of items to prevent bias. Importantly, we use item response models to calibrate the data. This means we can accurately measure item difficulty, taking into considering participant ability, as well as discirmination and guessing. This in turn can be used to construct new tests with known properties, by test information or test functioning. We use this to construct three new short forms, which we test. We make all data and parameters public.y

ALL LINKS TO REPOS SHOULD BE HERE

@@

A total of N=1584 participants were recruited from the Prolific Academic platform (https://www.prolific.co) to participate in an online behavioral experiment in July - August, 2021. Participants were eligible if they currently resided in the United States. This study was approved by the Institutional Review Board of Princeton University (\#7392), and all participants provided informed consent. Total study duration was approximately 6.4 minutes (2.4 sd) per participant. Participants received monetary compensation for their time (rate USD \$10/hr), plus an incentive-compatible bonus up to \$0.50 based on task performance (average payment: mean = \$1.30 [\$0.10 sd]). The experiment

cite https://doi.org/10.1111/bjop.12288, https://doi.org/10.1073/pnas.1018601108

After providing consent, participants completed a series of items from the matrix reasoning item bank. The design of the MaRs-IB has been described elsewhere \citep{chierchia2019matrix} and was based on publicly available code (https://app.gorilla.sc/openmaterials/36164). Briefly, each MaRs-IB item consisted of a 3 × 3 matrix. Eight of the nine resulting cells contained an abstract shape, while one cell on the bottom right-hand side of the matrix was empty. The participants' task was to complete the matrix by finding the missing shape among four possible alternatives. Items began with a 1200 ms fixation cross. Participants were then given up to 30 s to complete an item. After 25 s a clock appeared, indicating that 5 s remained before the next item began. An item ended when participants responded, or after 30 s had elapsed without response. Participants reviewed instructions before the task and were made to correctly complete three practice items before they could begin the task.

In order to ensure adequate sampling of all items, and in contrast to the original task design, all participants completed a pseudo-randomly selected subset of 16 out of 64 possible MaRs-IB items \footnote{We did not collect new data on 16 of the original MaRs-IB items. We excluded all Dimension 1 items, which are the simplest items and tend to have performance levels at ceiling. We also excluded Item 1 which, although not Dimension 1, also has performance levels near ceiling.} Item selection proceeded as follows: The 64 items were sorted into 16 groups by dimension. Participants were then randomly assigned one item from each of the groups. Importantly, participants were randomly assigned a clone of each item that varied in distractor type (md, pd) and item set (1, 2, 3). In this way, all items were sampled roughly 375 times (12 sd) and each clone was sampled 62 times (7 sd).

The data from multiple participants who completed the experiment were excluded prior to analysis. The data from N=7 participants were excluded for having browser resolutions smaller than necessary to see the task. The data from N=8 participants were excluded for missing responses on 4 or more trials. The data from N=70 participants were excluded for rapid guessing (defined as response times of less than 3 s) on 4 or more trials. In total, 83 of 1584 (5.24\%) were excluded, leaving the data from N=1501 participants for analysis.

The majority of participants in the calibration sample identified as women (men: N=670; women: N=811; non-binary or other: N=13; rather not say: N=2). Participants were 28.7 years old (9.9 sd) on average. The sample was relatively well-educated with the majority having a Bachelor's degree (N=507) or Master's degree or higher (N=322); fewer participants endorsed only completing some college (N=471) or only a high school degree (N=199).

The experiment was programmed in jsPsych \citep{de2015jspsych} and distributed using custom web-application software. The experiment code is available at \url{https://github.com/szorowi1/mars-irt}, and the web-software is available at \url{https://github.com/nivlab/nivturk}. A playable demo of the task is available at \url{https://nivlab.github.io/jspsych-demos/tasks/mars/experiment.html}.

@@

On average, participants completed 9.6 of 16 items correctly (sd = 3.1, IQR = 8.0 - 12.0) (Figure \ref{fig:fig01}a). Timing out occurred in only 1.8\% of items. A total of 43 participants (2.9\%) performed below chance (i.e. fewer than four items correct), while only 11 participants (0.7\%) responded correctly to all items. Thus, over 95\% of participants' observed scores were within a reasonable range. These results corroborate \cite{chierchia2019matrix} in that there were no obvious ceiling or floor effects, and that participants were motivated to perform despite minimal supervision in the online environment.

Across all 384 items, the average proportion correct was 0.60 (sd = 0.19, IQR = 0.45 - 0.74) (Figure \ref{fig:fig01}b). A total of 12 items (3.1\%) had performance beneath chance levels (25\%), and only 22 items (5.7\%) had performance levels above 90\%. As such, the majority of items sampled exhibited adequate performance with no obvious ceiling or floor effects.

Though not the primary focus of the analyses, we briefly investigated the relationship between participant' total scores, item difficulty, and response times. We fit a mixed-effects linear model to regress the log response time for trial $i$ on accuracy, the participant's observed score (minus their accuracy on that trial), the difficulty of the time (one minus the proportion correct), and the interaction of observed score and difficulty. We specified an intercept per participant to control for each participants' overall speed.

On average, participants spent 15.9 seconds to solve an item (sd = 7.2 s, IQR = 10.1 - 21.5 s). The mixed-effects model identified a significant effect of accuracy ($\beta = 0.067, t = 9.84, p < 1e^{-6}$), such that correct responses were on average slower than incorrect responses. Notably, there were large, significant effects of observed score ($\beta = 0.125, t = 16.948, p < 1e^{-6}$) and item difficulty ($\beta = 0.137, t = 44.326, p < 1e^{-6}$) such that participants were slower overall for more difficult items and for better performing participants. Finally, the interaction between observed score and item difficulty was also significant ($\beta = 0.058, t = 20.650, p < 1e^{-6}$), such that better performing participants slowed down even more for more difficult items.

The full pattern of response time results is summarized in Figure \ref{fig:fig01}c. In short, we corroborate the previous finding that item difficulty is associated with slower response times \citep{chierchia2019matrix}. In addition, we find two interesting patterns: better performing participants are slower on average, and this slowing is enhanced as items become more difficult. What this suggests then is that participants' overall performance on the MaRs-IB in this sample is reflective of speed-accuracy trade-offs likely reflective of individual differences in persistence or willingness to expend mental effort \citep{ranger2021effects}. That is, better performing participants are willing to spend more time on an item, whereas worse performing participants are more likely to give up sooner and guess as items become more difficult.

@@

@@

Reflecting the inherent hierarchical structure of the MaRs-IB -- 80 item families with six clones each -- we estimated the psychometric properties of each item using item cloning models \citep{geerlings2011modeling, cho2014additive, lathrop2017item}. All models were extensions of the three-parameter logistic (3PL) item repsonse model, where the probability of a correct response ($y_{ijk} = 1$) for person $i$ on item clone $k$ belonging to item family $j$ is:

@@

\noindent where $\theta_i$ is the latent ability for person $i$, and $\beta_{jk}$, $\alpha_{jk}$, and $\gamma_{jk}$ are the difficulty, discrimination, and guessing parameters for item clone $k$ of item family $j$. Because guessing parameters are difficult to estimate without very large amounts of data \citep{han2012fixing}, we fixed the guessing rate for each item to the nominal rate ($\gamma_{jk} = 0.25$).

@@

\noindent The difficulty parameter is thus modeled as an intercept ($\mu_\beta$), reflecting the average difficulty across all items, and four components:

@@

\item $\sum_{n=1}^N Q_{jn} \delta_{\beta n}$: the effect of the level-1 attributes, where $Q_{jn}$ is the value of attribute $n$ for item family $j$, and $\delta_{\beta n}$ is the effect of attribute $n$. This component is the fixed effects contribution of item family attributes to item difficulty.

\item $\epsilon_{\beta j}$: the level-1 residual, or variance unexplained by the item family attributes. This component is the random effects contribution of item family attributes to item difficulty.

\item $\sum_{m=1}^M R_{km} \delta_{\beta m}$: the effect of the level-2 attributes, where $R_{km}$ is the value of attribute $m$ for item clone $k$, and $\delta_{\beta m}$ is the effect of attribute $m$. This component is the fixed effects contribution of item clone attributes to item difficulty.

\item $\epsilon_{\beta k}$: the level-2 residual, or variance unexplained by the item clone attributes. This component is the random effects contribution of item clone attributes to item difficulty.

@@

To elaborate, level-1 attributes describe the properties of an item family. Here we model the effects of two properties: feature number, or the number of unique shapes in each cell of an item; and rule number, or the number of changing relationships among the shapes (e.g. color change, shape change, position change) one has to consider in order to correctly solve the item. Example items illustrating these attributes are included in the Supplement.  Note that this definition of item rules deviates from other established taxonomies of matrix reasoning items \citep{carpenter1990one}, a point we return to in the Discussion.

In turn, level-2 attributes describe properties of an item clone. Here we model the effects of two additional properties: distractor type, or whether the distractors were generated following the minimal difference (MD) strategy and a paired difference (PD) strategy (see \cite{chierchia2019matrix} for details); and mean response time, or the average time it took to complete that item. The latter attribute was orthogonalized with respect to the three other item attributes and thus reflects any residual structure after accounting for all other modeled effects.

@@

\noindent The fixed effect components of the item discrimination parameters were estimated using the same level-1 and level-2 attributes as for the item difficulty parameters (i.e. number of features, number of rules, distractor type, average response time). During estimation, the item discrimination parameters were restricted to be in the range $\alpha_{jk} \in [0, 5]$.

Though complex, the item structure model above provides a natural way of measuring the quality of the MaRs-IB. In the perfect situation, the clones of each item family would be psychometrically equivalent. So too should item families with identical attributes (e.g. number of features, number of rules). This is, of course, unlikely in actuality and the item cloning model allows us to quantify the extent to which equivalent families and clones vary in their difficulty and discrimination.

To quantify the success of the item cloning process, we estimated a series of four models \citep{cho2014additive, lathrop2017item}:

@@

\item \textit{Model 1}: Item parameters are expressed only as a function of their level-1 attributes. This model treats all clones within a family as interchangeable and all item families with the same attributes as equivalent.

\item \textit{Model 2}: Item parameters are expressed as a function of their level-1 attributes and residuals. This model treats all clones within a family as interchangeable, but now item families with the same attributes may psychometrically vary.

\item \textit{Model 3}: Item parameters are expressed as a function of their level-1 attributes, level-2 attributes, and level-1 residuals. This model allows clones within a family to vary, but all variation must be explained solely through their attributes.

\item \textit{Model 4}: Item parameters are expressed as a function of their level-1/2 attributes residuals. The full model in which item families and clones may vary in their psychometric properties.

@@

The resulting models were then submitted to model comparison. This procedure allows us to measure the success of the item generation process. If the psychometric properties of item families reflect only their attributes, then Model 1 should be preferred to Model 2; in contrast, if item families vary in their properties above and beyond what their attributes can explain, then Model 2 should be preferred. Similarly, if the clones of item family are truly equivalent, then Models 1 or 2 should be preferred to Models 3 or 4; in contrast, if item clones vary substantially (systematically or otherwise), then Models 3 or 4 should be preferred to Models 1 or 2.

Finally, we estimated person-level abilities following an item response modeling approach \citep{wilson2008explanatory}:

@@

\noindent where $X_{ip}$ is the value of attribute $p$ for person $i$, $\rho_p$ is the effect of attribute $n$, and $\epsilon_i$ is variance in ability unexplained by the person-level attributes.

We modeled person-level abilities as a function of four attributes: age, gender (coded as: male = -0.5, female = 0.5; all else = 0), average response time, and $\Delta$ response time. The last term reflected a person's adjustment of their response time as a function of item difficulty (defined as its corresponding proportion of correct responses), and thus measured a participant's tendency to slow down (or speed up) in response to more challenging items.

To ensure the identifiability of all models, person abilities were assumed to be $\theta_i \sim \mathcal{N} \left(0,1\right)$.

\begin{equation}
blah 
\end{equation}

All models were estimated within a Bayesian framework using Hamiltonian Monte Carlo as implemented in Stan (v2.26) \citep{carpenter2017stan}. For all models, four separate chains with randomised start values each took 7500 samples from the posterior. The first 5000 samples from each chain were discarded. As a result, 10000 post-warmup samples from the joint posterior were retained. The $\hat{R}$ values for all parameters was less than 1.01, indicating acceptable convergence between chains, and there were no divergent transitions in any chain.

Models were compared using Bayesian leave-one out cross-validation \citep{vehtari2017practical}, which has been found to perform better than other information criteria for comparing item response models \citep{luo2017performances}. Importantly, we computed the marginalized leave-one-cluster out (LOCO) cross-validation \citep{merkle2019bayesian}, which measures models' ability to generalize to held out items (rather than held out responses).

The fit of the winning model to the data was evaluated through several approaches. First, the fit of the models to the data was assessed by means of posterior predictive model checking \citep{gelman1996posterior}. Item fit was assessed using two two discrepancy statistics: Orlando and Thissen index ($S-\chi^2$) \citep{toribio2011discrepancy} and the generalized dimensionality discrepancy measure (GDDM) \citep{levy2009posterior, levy2011generalized}. The $S-\chi^2$ index is a measure of item fit based on the comparison of observed and expected proportions of participants who answer an item correctly at each raw-score group. The GDDM is a measure to test the model's unidimensionality assumption, which must not be violated in order to compute valid total scores from tests assembled from MaRs-IB items.

@@

First we estimated Models 1-4 holding constant the item discrimination parameters ($\alpha = 1$), which allowed us to investigate the difficulty of MaRs-IB items independently. (Note that these models still included guessing parameters fixed to the nominal guessing rate. We also estimated models with no guessing parameter, i.e. 1PL or Rasch models, but model comparison revealed these to be inferior.) The goodness of fit of the four models was compared based on leave-one-cluster out (LOCO) cross-validation as summarized in Table \ref{table:1}.

@@

On average, items were difficult (95\% HDI), but ranged considerably in difficulty. Over half of the variability in item difficulty (65.6\%) was explained by the fixed effects of item family (level-1) and clone (level-2) attributes. The associations between level-1 attributes and item difficulty were as expected: feature number was positively associated with difficulty (beta), such that the inclusion of one additional feature was associated with a 13.1\% reduction in expected accuracy. Similarly, rule number was positively associated with item difficulty (beta), such that the inclusion of one additional rule a 10.1\% reduction in expected accuracy. Turning next to level-2 attributes, distractor type was unexpectedly associated with item difficulty (beta), such that clones with MD distractors clones were associated with a 13.6\% reduction in accuracy compared to clones with PD distractors. This pattern is clearly evident in the raw item-level scores (Figure XX). Finally, average response time was positively associated with difficulty, such that a one standard deviation change in response time was associated with an 11.8\% reduction in accuracy.

@@

Paragraph XX: model more features as future direction. Carpenter taxonomy. Model by rule type. By perceptual features. automatic item generation as future direction. See \cite{lathrop2017item} for inspiration.

\bibliography{example}

\begin{table}
\centering
\begin{tabular}{l|c|c|c}
Model & elpd$_{loco}$ & loco & $\Delta$ loco \\\hline
1PG-F &  & -14516 & \\
1PG-FX &  & -13804 & \\
1PG-R &  & -13481 & \\
1PG-RX &  & -13293 & - \\
\end{tabular}
\caption{Table to test captions and labels.}
\label{table:1}
\end{table}

>>>>

Despite their utility, there are a number of challenges in using matrix reasoning measures in behavioral research. The first is the issue of copyright. The most prominent versions, including the Raven's progressive matrices \citep{raven2003raven} and the matrix subtest of the Wechsler Abbreviated Scale of Intelligence \citep{wechsler1999wechsler}, are not free to use and copyright may prevent these pen-and-paper tasks from being adapted into computerized tasks. A second challenge is those measures that are in the public domain typically have few items available. For example, the Hagen Matrices Test \citep{heydasch2014hagen} and International Cognitive Ability Resource \citep{condon2014international} have only 20 and 16 items available, respectively. The availability of only a limited number of items raises the issue of practice effects \citep{ng1974applicability, bors2003effect}, especially in the context of online experiments where multiple groups of researchers may be unknowingly administering the same items to the same participants.

In response to these challenges, \cite{chierchia2019matrix} developed the matrix reasoning item bank (MaRs-IB). The MaRs-IB is an open-source, free-to-use bank of 80 matrix reasoning items, where each item has three unique shape variants and two sets of distractors. The items were designed to vary considerably in their complexity (see Figure 1 for examples). In an initial validation study of a MaRs-IB short-form test, \cite{chierchia2019matrix} that measures of performance in a sample of 659 adults, adolescents, and children had excellent split-half and test-retest reliability. Moreover, performance on the MaRs-IB items were found to correlate with performance on the digit span tasks and ICAR matrix reasoning, indicating good convergent validity. Most importantly, the authors made available the percent correct data for each item (and variant) so that others researchers would be able to construct new MaRs-IB forms of custom difficulty and duration.

There are unfortunately two key obstacles preventing researchers from being fully able to do so, both of which stem from the design of the validation study. The original MaRs-IB short-form is a fixed-time and  fixed-order measure, where participants complete as many items as possible within 8-minutes. Because participants only completed 33 items on average, the first issue is that most items were completed by too few participants to establish reliable norms (an issue that is compounded when considering each item has six variants). Just over half the items (42/80) were completed by 100 or more participants. A second issue is that, because participants were free to choose for themselves whether to prioritize accuracy (number of correct responses) or productivity (number of items reached), it is likely that the percent correct data for items reached by fewer numbers of participants are biased. That is, items completed by fewer numbers of participants are more likely to have been completed by participants sacrificing accuracy for speed. Indeed, looking only at the easiest MaRs-IB items (dimension = 1), there is a strong positive correlation between number of completing participants and mean accuracy ($r = 0.673$).

Here we wish to overcome the limitations of the previous study to further validate and make useful the MaRs-IB. First, we collected a novel dataset of response to MaRs-IB items in a large sample of adults (N=1501). Using a fixed-length design with counterbalancing, we ensured that all items, and their variants, were completed by a suitable number of participants. Next we analyzed participants' response data hierarchical item response models, which allowed us to measure all items' difficulty and discrimination while controlling for participants' varying abilities. The item response models also allowed us to measure directly the extent to which that item variants are truly equivalent. We then demonstrate how these item response parameters can be used to construct new short-forms with desirable properties. Using these methods, we introduce three new MaRs-IB short-form tests with good internal consistency and convergent validity. 
All data and code, including items' estimated parameters and example code for test assembly, is available at: \url{https://github.com/szorowi1/mars-irt}.

@@

A total of N=1584 participants were recruited from the Prolific Academic platform (https://www.prolific.co) to participate in an online behavioral experiment in July - August, 2021. Participants were eligible if they were 18 years or older and currently resided in the United States. This study was approved by the Institutional Review Board of Princeton University (\#7392), and all participants provided informed consent. Total study duration was approximately 6.4 minutes (2.4 sd) per participant. Participants received monetary compensation for their time (rate USD \$10/hr), plus an incentive-compatible bonus up to \$0.50 based on task performance (average payment: mean = \$1.30 [\$0.10 sd]). Monetary incentives have previously been shown to motivate good performance in low-stakes testing environments \citep{duckworth2011role, gignac2018moderate}.

After providing consent, participants completed a series of items from the MaRs-IB. The design of the MaRs-IB items have been described previously \citep{chierchia2019matrix}. Briefly, each MaRs-IB item consists of a 3 x 3 matrix. Eight of the nine resulting cells contain an abstract shape, while one cell on the bottom right-hand side of the matrix is empty. The participants' task was to complete the matrix by finding the missing shape among four possible alternatives. Items began with a 1200 ms fixation cross. Participants were then given up to 30 s to complete an item. After 25 s a clock appeared, indicating that 5 s remained before the next item began. An item ended when participants responded, or after 30 s had elapsed without response. Participants reviewed instructions before the task and were made to correctly complete three practice items before they could begin the task. Crucially, participants were instructed to respond carefully but to guess before the timer ran out. Our version of the MaRs-IB was based on publicly available code (\url{https://app.gorilla.sc/openmaterials/36164}).

In order to ensure adequate sampling of all items, and in contrast to the original MaRs-IB test form, all participants completed a pseudorandomly-selected set of 16 out of 64 possible MaRs-IB items \footnote{We did not collect new data for the 16 easiest items from the MaRs-IB (items 1, 2, 3, 4, 5, 7, 8, 9, 32, 33, 38, 41, 43, 48, 57, 68) as the original study or pilot testing found performance on these items to be at or near ceiling.}. Prior to data collection, we sorted the 64 items into 16 groups of four based on the number of changing relationships between the shapes of the matrix. Participants were then randomly assigned one item from each of the 16 groups. In this way, all participants received test forms of approximately equal difficulty. Importantly we also took steps to counterbalance the sampling of item variants, so that we had approximately the same number of responses available by item set (1, 2, or 3) and distractor type (minimal difference or or paired difference; see \cite{chierchia2019matrix} for details).

To ensure data quality, the data from multiple participants who completed the experiment were excluded prior to analysis for one or more of the following reasons: for rapid guessing -- defined here as responding in less than 3 s -- on 4 or more trials N=70); for failing to respond on four or more trials (N=8); for minimizing or otherwise shrinking the browser window beyond what was required to view the stimuli (N=7). In total, 83 of 1584 (5.24\%) participants were excluded, leaving the data from N=1501 participants for analysis. Among these participants, each item (64 in total) was sampled approximately 375 times (sd = 12) and each item variant (384 in total) was sampled approximately 62 times (sd = 7).

@@

The majority of participants in the sample identified as women (men: N=670; women: N=811; non-binary or other: N=13; rather not say: N=2) and were 28.7 years old (sd = 9.9) on average. The sample was relatively well-educated, with the majority having completed a Bachelor's degree (N=507) or Master's degree or higher (N=322). By comparison, fewer participants endorsed having completed only some college (N=471), only a high school degree (N=199), or preferring not to say (N=2).

On average, participants completed 9.6 of 16 items correctly (sd = 3.1, IQR = 8.0 - 12.0) (Figure \ref{fig:fig01}a). Timing out occurred on only 1.8\% of trials. A total of 43 participants (2.9\%) performed below chance (i.e. fewer than four items correct), while only 11 participants (0.7\%) correctly responded to all 16 items. Thus, over 95\% of participants' observed scores were within a reasonable range. These results corroborate \cite{chierchia2019matrix} in that there were no obvious ceiling or floor effects in performance, and that the majority of participants were sufficiently motivated to participate despite the low-stakes testing environment.

Across all 384 unique item variants, the average proportion correct was 0.60 (sd = 0.19, IQR = 0.45 - 0.74) (Figure \ref{fig:fig01}b). A total of 12 items (3.1\%) had accuracy rates beneath chance levels (25\%), and only 22 items (5.7\%) had performance levels above 90\%. Mean performance on an item was significantly and negatively correlated with the number of changing relationships amongst its features (Spearman's $\rho = -0.482, p < 1e^{-6})$). Moreover, item-wise accuracy rates in this sample were remarkably similar to those reported by \cite{chierchia2019matrix} (Spearman's $\rho = 0.870, p < 1e^{-6})$), when restricting the latter to items with more than 200 available responses. These results further corroborate \cite{chierchia2019matrix} in that the majority of items exhibited good performance, with few items at ceiling or floor and passing of the manipulation check.

Though not the primary focus of the current study, we briefly investigated the relationship between response accuracy and times. We used a mixed-effects model to predict participants' trial-level log-transformed response times as a function their response on that trial (correct or incorrect), their rest score (their total score on all other items), that item's difficulty (one minus the accuracy rate), and all possible interactions. We also specified an intercept per participant to control for each participants' overall speed.

The complete summary of results from the regression analysis are presented in Table XX. On average, participants spent 15.9 s per item (sd = 7.2 s, IQR = 10.1 - 21.5 s). The mixed-effects model identified significant main effects of accuracy ($\beta = 0.067, t = 9.84, p < 1e^{-6}$), rest score ($\beta = 0.125, t = 16.948, p < 1e^{-6}$), and item difficulty ($\beta = 0.137, t = 44.326, p < 1e^{-6}$). On average then, response times were slower for correct responses, more accurate participants, and more difficult items. There was also a significant positive accuracy by difficulty interaction (correct responses were even slower for more difficult items) and a significant negative accuracy by score interaction (correct responses were faster for better performing participants). Finally, there was a significant positive difficulty by score interaction, such that better performing participants exhibited slower responses for more difficult items. The three-way interaction term was not statistically significant.

In short, we corroborate the previous finding that item difficulty is associated with slower response times \citep{chierchia2019matrix}. In addition, we identified two notable patterns: better performing participants exhibited slower response times on average, which slowered further for more difficult times. This pattern of response time results is summarized in Figure \ref{fig:fig01}c. This pattern of results is consistent with controlled processing accounts, a point we return to in the Discussion.

@@

Summary paragraph?

@@

To measure the psychometric properties of each item, we used item response models. Given the hierarchical nature of the MaRs-IB -- 64 items with six variants each -- we specifically modeled participants' response data using item using a series of item cloning models \citep{geerlings2011modeling, cho2014additive, lathrop2017item}. In item cloning models, the variants of an item (or "clones") are explicitly model as having been generated from the same template item (or "family"). The basis of all of the models used here is the three-parameter logistic (3PL) item response model, where the probability of a correct response ($y_{ijk} = 1$) for person $i$ on item clone $k$ belonging to item family $j$ is:

@@

\noindent where $\theta_i$ is the latent ability for person $i$, and $\beta_{jk}$, $\alpha_{jk}$, and $\gamma_{jk}$ are the difficulty, discrimination, and guessing parameters for item clone $k$ of item family $j$. Since guessing parameters are challenging to estimate in the absence of very large amounts of data \citep{han2012fixing}, we fixed the guessing parameter for each item clone to the nominal guessing rate ($\gamma_{jk} = 0.25$).

@@

\noindent Every difficulty parameter is thus modeled as an intercept ($\mu_\beta$), reflecting the average difficulty across all items, and four components:

@@

\item $\sum_{n=1}^N Q_{jn} \delta_{\beta n}$: the effect of the level-1 attributes, where $Q_{jn}$ is the value of attribute $n$ for item family $j$, and $\delta_{\beta n}$ is the effect of attribute $n$. This component is the fixed effects contribution of item family to item difficulty.

\item $\epsilon_{\beta j}$: the level-1 residual, or variance unexplained by the item family attributes. This component is the random effects contribution of item family to item difficulty.

\item $\sum_{m=1}^M R_{km} \delta_{\beta m}$: the effect of the level-2 attributes, where $R_{km}$ is the value of attribute $m$ for item clone $k$, and $\delta_{\beta m}$ is the effect of attribute $m$. This component is the fixed effects contribution of item clone to item difficulty.

\item $\epsilon_{\beta k}$: the level-2 residual, or variance unexplained by the item clone attributes. This component is the random effects contribution of item clone to item difficulty.

@@

To elaborate, level-1 attributes describe the properties of an item family. Here we model the effects of two properties: feature number, or the number of unique shapes in each cell of an item; and rule number, or the number of changing relationships among the shapes (e.g. color change, shape change, position change) a participant must consider in order to solve the item.

In turn, level-2 attributes describe properties of an item clone. Here we model the effects of two additional properties: distractor type, or whether the distractors were generated following the minimal difference (MD) strategy and a paired difference (PD) strategy; and mean response time, or the average time spent on that item. Mean response time was orthogonalized with respect to the three other item attributes, and therefore reflects any residual structure after accounting for all other modeled effects.

@@

\noindent The fixed effect components of the item discrimination parameters included the same level-1 and level-2 attributes as for the item difficulty parameters (i.e. number of features, number of rules, distractor type, average response time). During estimation, the item discrimination parameters were restricted to be in the range $\alpha_{jk} \in [0, 5]$.

Though complex, item cloning models provide a natural means of evaluating the success of the MaRs-IB item generation process. For example, modeling an item's properties as a function of its attributes allows us to quantify the extent to which item design worked as intended (e.g. that an increase in the number of changing relationships between shapes corresponds to an increase in item difficulty). The inclusion of level-1 random effects allows us to quantify the extent to which item families with identical attributes (e.g. number of features, number of changing relations) are psychometrically equivalent. Similarly, the inclusion of level-2 random effects allows us to quantify the extent to which clones of an item family are psychometrically equivalent.

In order to test the hypotheses outlined above, we fit a series of nested item cloning models \citep{cho2014additive, lathrop2017item}:

@@

\item \textit{Model a}: Item parameters are expressed only as a function of their level-1 attributes. This model treats all clones within a family as interchangeable and all item families with the same attributes as equivalent.

\item \textit{Model b}: Item parameters are expressed as a function of their level-1 attributes and residuals. This model treats all clones within a family as interchangeable, but now item families with the same attributes may vary in their properties.

\item \textit{Model c}: Item parameters are expressed as a function of their level-1 attributes, level-2 attributes, and level-1 residuals. This model allows clones within a family to vary, but all variation must be explained solely through their attributes.

\item \textit{Model d}: Item parameters are expressed as a function of their level-1/2 attributes residuals. The full model in which item families and clones may vary in their psychometric properties.

@@

The resulting model fits are then compared as a test of the hypotheses above. For example, if the psychometric properties of items can be solely explained by their attributes, then Model A should be preferred to B; in contrast, if item families psychometrically vary beyond what their attributes can explain, then Model B should be preferred. Similarly, if the clones of an item family are truly equivalent, then Models A or B should be preferred; in contrast, if item clones vary (as a function of their attributes or otherwise) then model C or D should be preferred.

Finally, we estimated person-level abilities following an explanatory item response modeling approach \citep{wilson2008explanatory}:

@@

\noindent where $X_{ip}$ is the value of attribute $p$ for person $i$, $\rho_p$ is the partial correlation of attribute $n$, and $\epsilon_i$ is variance in ability unexplained by the person-level attributes.

We modeled person-level abilities as a function of four attributes: age, gender (coded as: male = -0.5, female = 0.5; all else = 0), average response time, and $\Delta$ response time. The last term reflected the degree to which a participant adjusted their response time as a function of item difficulty, and thus measured a participant's tendency to slow down (or speed up) in response to more challenging items (Figure \ref{fig:fig01}c).

To ensure the identifiability of all models, person abilities were constrained to have a mean of zero and a variance of 1.0, such that the residual variance was defined as $V(\epsilon) = 1 - \sum \rho_p^2$. Note that all person-level attributes were standard-scored prior being entered into the model.

All models were estimated within a Bayesian framework using Hamiltonian Monte Carlo as implemented in Stan (v2.26) \citep{carpenter2017stan}. For all models, four separate chains with randomised start values each took 7500 samples from the posterior. The first 5000 samples from each chain were discarded. As such, 10,000 post-warmup samples from the joint posterior were retained. The $\hat{R}$ values for all parameters was equal to or less than 1.01, indicating acceptable convergence between chains, and there were no divergent transitions in any chain.

Models were compared using Bayesian leave-one out cross-validation \citep{vehtari2017practical}, which has been found to perform better than more traditional information criteria for comparing item response models \citep{luo2017performances}. Importantly, we computed the conditional leave-one-cluster out (LOCO) cross- validation \citep{merkle2019bayesian}, which measures a model's ability to generalize to held out items (rather than held out responses).

The fit of the best-fitting model to the data was evaluated through posterior predictive model checks \citep{gelman1996posterior, levy2017bayesian}. Item fit was assessed using the Orlando and Thissen index ($S-\chi^2$) \citep{toribio2011discrepancy}, which compares the observed and expected proportions of participants who answer an item correctly at each raw-score group. The unidimensionality of the data was assessed using the generalized dimensionality discrepancy measure (GDDM) \citep{levy2011generalized}, which tests for violation of the assumption of local independence. This latter measure is crucial for ensuring there is not non-negligible levels of local item dependence in the data, which may otherwise lower the efficiency of future test forms due to the inclusion of correlated, redundant items.

@@

First we estimated Models 1-4 holding constant the item discrimination parameters ($\alpha = 1$), which allowed us to investigate the difficulty of MaRs-IB items independently. (Note that these models still included guessing parameters fixed to the nominal guessing rate. We also estimated models with no guessing parameter, i.e. 1PL or Rasch models, but model comparison revealed these to be inferior.)

The goodness of fit of the four models was compared based on leave-one-cluster out (LOCO) cross-validation as summarized in Table \ref{table:1}. The winning model was Model 1d, in which item difficulties were modeled as a function of all level-1 and level-2 fixed and random effects. This has multiple interpretations, which we will unpack in turn. First, this implies that there was systematic variability in item difficulty at the family-level beyond what could explained by their attributes (this is reflected...

[above to be combined with below when time comes]

@@

The average item difficulty was $\beta_\mu = 0.177$ (95\% HDI = 0.118 - 0.238) but this ranged considerably across items (sd = 1.431, 95\% HDI = 1.370 - 1.492).

Over half of the variability in item difficulty (65.6\%) was explained by the fixed effects of item family (level-1) and clone (level-2) attributes. The associations between level-1 attributes and item difficulty were as expected: feature number was positively associated with difficulty (beta), such that the inclusion of one additional feature was associated with a 13.1\% reduction in expected accuracy. Similarly, rule number was positively associated with item difficulty (beta), such that the inclusion of one additional rule a 10.1\% reduction in expected accuracy. Turning next to level-2 attributes, distractor type was unexpectedly associated with item difficulty (beta), such that clones with MD distractors clones were associated with a 13.6\% reduction in accuracy compared to clones with PD distractors. This pattern is clearly evident in the raw item-level scores (Figure XX). Finally, average response time was positively associated with difficulty, such that a one standard deviation change in response time was associated with an 11.8\% reduction in accuracy.

@@

Paragraph XX: model more features as future direction. Carpenter taxonomy. Model by rule type. By perceptual features. automatic item generation as future direction. See \cite{lathrop2017item} for inspiration. Note that this definition of item rules deviates from other established taxonomies of matrix reasoning items \citep{carpenter1990one}, a point we return to in the Discussion.

\bibliography{references}
